UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-11-16 ms/main.tex
minor tweaks

\begin{abstract}
%Max 250
Background: Genetic structure in large cohorts can arise from technical, sampling and demographic variation. Data visualisation is therefore a first step in most genomic analyses. However, existing data exploration methods struggle with unbalanced sampling and the multiple scales of population structure in large cohorts. We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
\\
\\
Results: Using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank, we demonstrate that PCA-UMAP effectively clusters individuals who are genetically closely related while placing them in a global continuum of genetic variation. This dimension reduction approach reveals previously overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank. 
\\
\\
Conclusion: PCA-UMAP reveals genetic subgroups among study participants that had been previously overlooked, opening new lines of investigation for demographic research and statistical genetics. Given its small computational cost, it provides an important general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.6\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \end{subfigure}
    \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six relevant 1KGP populations are presented: CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
\end{figure*}

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.pdf}%
\caption{Principal components 1 and 2}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{UMAP on first 10 principal components}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{The UKBB projected onto two dimensions, colored by self-reported ethnic background. (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. (b) UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
%The main body of the image is made up of British and Irish individuals. Section (i) is zoomed in on the structure of Black/Black British individuals. Section (ii) is focused on East and South Asian populations, and section (iii) is a mix of a variety of ethnicities.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest body in the figure consists of the White British and Irish populations. The Irish population concentrates in a portion of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within what is suspected to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals. These trails come together in a nexus of individuals with a variety of ethnicities; many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group". Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asia backgrounds. Additionally, there are many individuals whose ethnicity is White but neither British nor Irish (AWB) forming clusters distinct from the British and Irish cluster.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the projection in figure~\ref{fig:umap_ukbb_pc10} colored in by  geographical coordinates according to the Ordnance Survey National Grid (OSGB1936) as distances north and east of the Isles of Scilly. UMAP coordinates within the "White British" cluster broadly map to geographic coordinates, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines connect to the South East corner of this cluster, corresponding to the position of the city of London and reflecting its high migrant population.

\begin{figure}
    \centering
    \includegraphics[width=0.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{The UKBB projected onto two dimensions using PCA-UMAP with each individual colored by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colors individuals by their north-south ("northing") coordinates, and the right image colors them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Individuals with missing geographic data are not shown. To prevent outlying individuals from washing out the colour scheme, northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. To protect participant privacy, data has been randomized as explained in the materials and methods section.}
    \label{fig:umap_ukbb_geo_osgb}
\end{figure}

@@

Covariates such as height (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures  (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure. Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and several unidentified ethnic groups have noticeably lower-than-average heights. More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots). Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.

 Forced expiratory volume in 1 second (FEV1) (Figure~\ref{fig:supp_box_fev_f} also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture continuum, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; while another close to European and British populations has lower values relative to the population mean. These results merit further investigation and underscore the exploratory value of PCA-UMAP --- these  populations are largely unidentified and there is no straightforward way to separate them within the data otherwise.

@@

%PCA-UMAP reveals numerous clusters that are based on shared ancestries, admixture proportion, and geography. Examples include the closeness of the Chinese Dai and Kinh of Vietnam and the diversity and admixture of the American Hispanic population.  Using the 1KGP, we illustrate in figures~\ref{fig:supp_megamontage_pc2_9} and~\ref{fig:supp_megamontage_pc10_50} that using progressively more principal components can create clusters on finer scales than previously demonstrated.

@@

Importantly, using UMAP is straightforward and fast. Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the a "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results). Given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.

@@

For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}

>>>>

\begin{abstract}
%Max 250
Background: Genetic structure in large cohorts can arise from technical, sampling and demographic variation. Data visualisation is therefore a first step in most genomic analyses. However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure in large cohorts. We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
\\
\\
Results: Using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank, we demonstrate that PCA-UMAP effectively clusters individuals who are genetically closely related while placing them in a global continuum of genetic variation. This dimension reduction approach reveals previously overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank. 
\\
\\
Conclusion: PCA-UMAP reveals genetic subgroups among study participants that had been previously overlooked, opening new lines of investigation for demographic research and statistical genetics. Given its small computational cost, it provides an important general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.6\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \end{subfigure}
    \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Coloring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions.}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
\end{figure*}

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.pdf}%
\caption{Principal components 1 and 2}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{UMAP on first 10 principal components}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{The UKBB projected onto two dimensions, colored by self-reported ethnic background. (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. (b) UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest body in the figure consists of the White British and Irish populations. The Irish population concentrates in a portion of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within appears to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals. These trails come together in a nexus of individuals with a variety of ethnicities; many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group". Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asian backgrounds. Additionally, there are many individuals whose ethnicity is White but neither British nor Irish (AWB) forming clusters distinct from the British and Irish cluster.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the projection in figure~\ref{fig:umap_ukbb_pc10} colored in by  geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. UMAP coordinates within the "White British" cluster broadly map to geographic coordinates, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines connect to the South East corner of this cluster, corresponding to the position of the city of London and reflecting its high migrant population.

\begin{figure}
    \centering
    \includegraphics[width=0.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{The UKBB projected onto two dimensions using PCA-UMAP with each individual colored by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colors individuals by their north-south ("northing") coordinates, and the right image colors them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Individuals with missing geographic data are not shown. To prevent outlying individuals from washing out the color scheme, northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. To protect participant privacy, data has been randomized as explained in the materials and methods section.}
    \label{fig:umap_ukbb_geo_osgb}
\end{figure}

@@

Covariates such as height (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures  (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure. Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and several unidentified ethnic groups have noticeably lower-than-average heights. More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t-test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots). Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture continuum, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; while another close to European and British populations has lower values relative to the population mean. These results merit further investigation and underscore the exploratory value of PCA-UMAP --- these  populations are largely unidentified and there is no straightforward way to separate them within the data otherwise.

@@

@@

Importantly, using UMAP is straightforward and fast. Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results). Given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.

@@

For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the color value from a randomly selected neighbour.}
